---
title: "Estimating Equations"
output:
  pagedown::html_paged:
    # change to true for a self-contained document, but it'll be a litte slower for Pandoc to render
    css: ["../style/my-style-page.css", "default-fonts", "default-page", "default"]
    self_contained: true
    number_sections: true
---

```{r, echo = FALSE, message = FALSE}
library(knitr)
library(ggplot2)
library(dplyr)
library(tidyr)
library(broom)
set.seed(333)

theme_set(theme_bw())
knitr::opts_chunk$set(fig.height = 3)
```

<br/><br/><br/><br/>

**Example:** Consider the $\boldsymbol Z = (Z_1, \dots, Z_5)^\top$ with cdf
$$
F(\boldsymbol z;\alpha) = \exp\left\{-\left(z_1^{-\frac{1}{\alpha}} + z_2^{-\frac{1}{\alpha}} + z_3^{-\frac{1}{\alpha}} + z_4^{-\frac{1}{\alpha}} + z_5^{-\frac{1}{\alpha}}\right)^\alpha\right\}, \quad \boldsymbol z \ge \boldsymbol 0, \alpha \in (0, 1].
$$

<br/><br/><br/><br/><br/><br/><br/><br/>

Comments:

1. $F$ is max-stable.

<br/><br/><br/><br/><br/><br/><br/><br/>

2. $Z_1, \dots, Z_5$ are exchangeable.

\newpage

Let's consider the likelihood.

\newpage

How about if we were to just use pairs of points to estimate $\alpha$?

<br/><br/><br/><br/><br/><br/><br/><br/><br/><br/><br/><br/><br/><br/><br/><br/>

If we just used $(z_{1i}, z_{2i}), i = 1, \dots, n$ would the likelihood based on the bivariate density be a good estimator for $\alpha$?

\newpage

Let's try it.

```{r, cache = TRUE}
library(evd)
# simulate data with alpha = 0.5
alpha <- 0.5
z <- rmvevd(500, dep = alpha, d = 5, mar = c(1, 1, 1))

## bivariate density
d_bivar <- function(z, alpha){
	#here "z" is a single observation (ordered pair)
	inside <- z[1]^(-1/alpha) + z[2]^(-1/alpha)
	one <- exp(-inside^alpha)
	two <- (z[1]*z[2])^(-1 / alpha - 1)
	three <- (1 / alpha - 1)*inside^(alpha - 2)
	four <- inside^(2 * alpha - 2)
	one*two*(three + four)
}
	
d_bivar(c(4, 5), alpha = alpha)
dmvevd(c(4,5), dep = alpha, d = 2, mar = c(1,1,1))

## estimate alpha
log_pair_lhood <- function(alpha, z) {
	#here "z" is bivariate matrix of observations
	inside <- z[, 1]^(-1 / alpha) + z[, 2]^(-1 / alpha)
	log_one <- -inside^alpha
	log_two <- (-1 / alpha - 1) * (log(z[, 1]) + log(z[, 2]))
	three <- (1 / alpha - 1) * inside^(alpha - 2)
	four <- inside^(2 * alpha - 2)	
	contrib <- log_one + log_two + log(three + four)
	return(sum(contrib))
}

all_pairs_lhood <- function(alpha, z) {
	expand.grid(dim1 = seq_len(ncol(z)),
	            dim2 = seq_len(ncol(z))) |>
	  filter(dim1 < dim2) |>
	  rowwise() |>
	  mutate(log_pair_lhood = log_pair_lhood(alpha, cbind(z[, dim1], z[, dim2]))) |>
    ungroup() |>
	  summarise(res = sum(log_pair_lhood)) |>
	  pull(res)
}

alpha_mple <- optim(.2, lower = .01, upper = .99, all_pairs_lhood, z = z, method = "Brent", hessian = TRUE, control = list(fnscale = -1))
(ci_mple <- alpha_mple$par + c(-1.96, 1.96)*sqrt(-1 / alpha_mple$hessian[1, 1]))

## checking coverage
#checking coverage
B <- 200
coverage <- numeric(B)
for(k in seq_len(B)) {
	z_k <- rmvevd(500, dep = .5, d = 5, mar = c(1, 1, 1))
	alpha_mple_k <- optim(.2, lower = .01, upper = .99, all_pairs_lhood, z = z_k, method = "Brent", hessian = TRUE, control = list(fnscale = -1))
	ci <- alpha_mple_k$par + c(-1.96, 1.96)*sqrt(-1 / alpha_mple_k$hessian[1, 1])
	coverage[k] <- as.numeric(ci[1] < alpha & ci[2] > alpha)
}
mean(coverage)
```

```{r, cache = TRUE, echo = FALSE, out.width="50%", fig.show='hold', fig.height=3.7}
log_lhood <- function(alpha, z) {
  sum(dmvevd(z, dep = alpha, d = 5, mar = c(1,1,1), log = TRUE))
}
alpha_mle <- optim(.2, lower = .01, upper = .99, log_lhood, z = z, method = "Brent", hessian = TRUE, control = list(fnscale = -1))
ci_mle <- alpha_mle$par + c(-1.96, 1.96)*sqrt(-1 / alpha_mle$hessian[1, 1])

data.frame(alpha_vec = seq(.2, .7, length.out = 500)) |>
  rowwise() |>
  mutate(actual_lhood = log_lhood(alpha_vec, z),
         pair_lhood = all_pairs_lhood(alpha_vec, z)) -> dat_plot

dat_plot |>
  ggplot() +
  geom_vline(aes(xintercept = alpha), lty = 2) +
  geom_line(aes(alpha_vec, pair_lhood), colour = "blue") +
  geom_segment(aes(x = ci_mple[1], xend = ci_mple[2], y = alpha_mple$value - 400, yend = alpha_mple$value - 400), colour = "blue", linewidth = 0.4) +
  geom_point(aes(x = alpha_mple$par, y = alpha_mple$value - 400), colour = "blue", size = 2) +
  ylim(c(min(c(dat_plot$actual_lhood, dat_plot$pair_lhood)), max(c(dat_plot$actual_lhood, dat_plot$pair_lhood)))) +
  ggtitle("Composite Log Likelihood") +
  xlab("alpha") +
  ylab("")


dat_plot |>
  ggplot() +
  geom_vline(aes(xintercept = alpha), lty = 2) +
  geom_line(aes(alpha_vec, actual_lhood), colour = "darkgreen") +
  geom_segment(aes(x = ci_mle[1], xend = ci_mle[2], y = alpha_mle$value - 400, yend = alpha_mle$value - 400), colour = "darkgreen", linewidth = 0.4) +
  geom_point(aes(x = alpha_mle$par, y = alpha_mle$value - 400), colour = "darkgreen", size = 2) +
  ylim(c(min(c(dat_plot$actual_lhood, dat_plot$pair_lhood)), max(c(dat_plot$actual_lhood, dat_plot$pair_lhood)))) +
  ggtitle("Actual Log Likelihood") +
  xlab("alpha") +
  ylab("")


```

\newpage

So, it looks like the point estimate from the pairwise likelihood is ok, but we need to be able to get an appropriate measure of uncertainty.

<br/><br/><br/><br/><br/><br/><br/><br/><br/><br/><br/><br/><br/><br/><br/><br/>

The proper adjustment is

\newpage

# Introduction

<br/><br/><br/><br/><br/><br/><br/><br/>

M-estimators are solutions of the vector equation
$$
\sum\limits_{i = 1}^n \boldsymbol \psi(\boldsymbol Y_i, \boldsymbol \theta) = \boldsymbol 0.
$$

<br/><br/><br/><br/><br/><br/><br/><br/>

In the likelihood setting, what is $\boldsymbol \psi$?

\newpage

**Example:** Let $Y_1, \dots, Y_n$ be independent, univariate random variables. Is $\theta = \overline{Y} = \frac{1}{n}\sum\limits_{i = 1}^n Y_i$ an M-estimator?

\newpage

**Example:** Consider the mean deviation from the sample mean,
$$
\hat{\theta}_1 = \frac{1}{n} \sum\limits_{i = 1}^n |Y_i - \overline{Y}|.
$$
Is this an M-estimator?

\newpage


# Basic Approach

M-estimators are solutions of the vector equation
$$
\sum\limits_{i = 1}^n \boldsymbol \psi(\boldsymbol Y_i, \boldsymbol \theta) = \boldsymbol 0.
$$
but what are they estimating?

<br/><br/><br/><br/><br/><br/>

**Example (Sample Mean, cont'd):** Recall we said $\theta = \overline{Y} = \frac{1}{n}\sum\limits_{i = 1}^n Y_i$ is an M-estimator for $\psi(Y_i, \theta) = Y_i - \theta$. What is the true parameter?

\newpage

To arrive at the sandwich estimator, assume $\boldsymbol Y_1, \dots, \boldsymbol Y_n \stackrel{iid}{\sim} F$ and define
$$
\boldsymbol G_n(\boldsymbol \theta) = \frac{1}{n}\sum\limits_{i = 1}^n \boldsymbol \psi(\boldsymbol Y_i; \boldsymbol \theta).
$$

<br/><br/><br/><br/><br/><br/>

Taylor expansion of $\boldsymbol G_n(\boldsymbol \theta)$ around $\boldsymbol \theta_0$ evaluated at $\hat{\boldsymbol \theta}$ yields

\newpage

Define $\boldsymbol A(\boldsymbol \theta_0) = \text{E}_F[-\boldsymbol \psi'(\boldsymbol Y_1, \boldsymbol \theta_0)]$.

\newpage

## Estimators for $\boldsymbol A, \boldsymbol B$

If the data truly come from the assumed parametric family $f(y; \boldsymbol \theta)$,

<br/><br/><br/><br/><br/><br/>

One of the key contributions of M-estimation theory is to point out what happens when the assumed parametric family is not correct.

<br/><br/><br/><br/>

We can use empirical estimators of $\boldsymbol A$ and $\boldsymbol B$:

\newpage

**Example (Coefficient of Variation):** Let $Y_1, \dots, Y_n$ be idd from some distribution with finite fourth moment. The coefficient of variation is defined at $\hat{\theta}_3 = s_n / \overline{Y}$.

<br/><br/><br/><br/><br/><br/>

Define a three dimensional $\boldsymbol \psi$ so that $\hat{\theta}_3$ is defined by summing the third component.  What is the vector valued function $\boldsymbol \psi$ which yields an M-estimator for the coefficient of variation?

\newpage

What parameter vector is being estimated by the M-estimator?


<br/><br/><br/><br/><br/><br/><br/><br/>

What are the matrices $\boldsymbol A$ and $\boldsymbol B$?

\newpage

Write out the asymptotic variance, $\boldsymbol V$.

\newpage

Assume $Y_i$ are iid from a normal distribution with mean $10$ and standard deviation $1$. Calculate $V_{3,3}$. Assume you have a same of size $25$ and you get an estimated coefficient of variation of $0.11$. Give the asymptotic $95$% confidence interval.

\newpage


