---
title: "Methods of Maximizing the Likelihood"
output:
  pagedown::html_paged:
    # change to true for a self-contained document, but it'll be a litte slower for Pandoc to render
    css: ["../style/my-style-page.css", "default-fonts", "default-page", "default"]
    self_contained: true
    number_sections: true
---

```{r, echo = FALSE, message = FALSE}
library(knitr)
library(ggplot2)
library(dplyr)
library(tidyr)
library(broom)
set.seed(333)

theme_set(theme_bw())
knitr::opts_chunk$set(fig.height = 3)
```

Maximum likelihood estimation requires maximization of the log likelihood $\ell(\boldsymbol \theta) = \log L(\boldsymbol \theta | \boldsymbol Y)$.



# EM Algorithm

Approach solving the likelihood equation via viewing the observed data $\boldsymbol Y$ as incomplete and that there is missing data $\boldsymbol Z$ that would make the problem simpler if we had it.

<br/><br/><br/><br/>

**Example (Two-Component Mixture):** Suppose $Y_1, \dots, Y_n$ are iid from the mixture density
$$
f(y; \boldsymbol \theta) = pf_1(y ; \boldsymbol \mu_1, \Sigma_1) + (1 - p)f_2(y ; \boldsymbol \mu_2, \Sigma_2),
$$
where $f_1$ and $f_2$ are bivariate normal densities with mean vectors $\boldsymbol \mu_1$ and $\boldsymbol \mu_2$ and variance matrices $\Sigma_1$ and $\Sigma_2$, respectively. Thus, the parameter vector $\boldsymbol \theta = (p, \boldsymbol \mu_1, \boldsymbol \mu_2, \Sigma_1, \Sigma_2)$ and the likelihood is
$$
L(p, \boldsymbol \mu_1, \boldsymbol \mu_2, \Sigma_1, \Sigma_2) = \qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad
$$

<br/><br/><br/><br/><br/><br/><br/><br/><br/><br/><br/><br/>
<br/><br/><br/><br/><br/><br/>

```{r, fig.show='hold', out.width="50%", fig.height=5}
library(mvtnorm) ## multivariate normal

p = .6
mu1 <- c(0, 0)
sig1 <- matrix(c(1, 0, 0, 1), ncol = 2)
mu2 <- c(1.5, 1.5)
sig2 <- matrix(c(1, .6, .6, 1), ncol = 2)

## sample from the mixture
n <- 50
z <- rbinom(n, 1, p)

y1 <- rmvnorm(sum(z), mean = mu1, sigma = sig1)
y2 <- rmvnorm(n - sum(z), mean = mu2, sigma = sig2)  
y <- matrix(NA, nrow = n, ncol = 2) ## observed data
y[z == 1, ] <- y1
y[z == 0, ] <- y2

df <- data.frame(y, z)

## plot data
ggplot(df) +
  geom_point(aes(X1, X2)) +
  ggtitle("Observed (Incomplete) Data")

ggplot(df) +
  geom_point(aes(X1, X2, colour = as.character(z))) +
  ggtitle("Complete Data")

```
Let's try to maximize the likelihood

```{r}
# loglikelihood of incomplete data--no knowledge of z
loglik_mixture <- function(par, data) {
	p <- plogis(par[1])  # p guaranteed to be in [0,1]
	mu1 <- c(par[2], par[3])
	sig1 <- matrix(c(exp(par[4]), par[5], par[5], exp(par[4])), nrow = 2)
	mu2 <- c(par[6], par[7])
	sig2 <- matrix(c(exp(par[8]), par[9], par[9], exp(par[8])), nrow = 2)
	# note:  exponential guarantees the diagonal elements are positive, but
	# nothing to guarantee matrices are positive definite. (Could do square root)

	out <- log(p * dmvnorm(data, mean = mu1, sigma = sig1) + 
	             (1-p) * dmvnorm(data, mean = mu2, sigma = sig2))
	return(sum(out))
}

## optimize from different starting values
mle1 <- optim(c(0, -.2, -.2, .5, 0, 2, 2, .5, 0), loglik_mixture, data = y, control = list(fnscale = -1))
mle2 <- optim(c(.405, 0, 0, 0, 0, 1.5, 1.5, 0, .6), loglik_mixture, data = y, control = list(fnscale = -1))
```

```{r, echo = FALSE}
data.frame(Parameter = c("$p$", "$\\mu_{11}$", "$\\mu_{12}$", "$\\Sigma_{111}$", "$\\Sigma_{112}$", "$\\mu_{21}$",  "$\\mu_{22}$", "$\\Sigma_{211}$", "$\\Sigma_{212}$"), 
           Truth = c(p, mu1, sig1[1:2], mu2, sig2[1:2]),
           MLE1 = c(plogis(mle1$par[1]), mle1$par[c(2,3)], exp(mle1$par[4]), mle1$par[5:7], exp(mle1$par[8]), mle1$par[9]),
           MLE2 = c(plogis(mle2$par[1]), mle2$par[c(2,3)], exp(mle1$par[4]), mle2$par[5:7], exp(mle2$par[8]), mle2$par[9])) |>
  kable(escape = FALSE, digits = 4)

```

<br/><br/><br/><br/>

Fitted results:

```{r, echo = FALSE, fig.show='hold', out.width="50%", fig.height=5}
hat_p <- plogis(mle2$par[1])
hat_mu1 <- mle2$par[2:3]
hat_sig1 <- matrix(c(exp(mle2$par[4]), mle2$par[5], mle2$par[5], exp(mle2$par[4])), nrow = 2)
hat_mu2 <- mle2$par[6:7]
hat_sig2 <- matrix(c(exp(mle2$par[8]), mle2$par[9], mle2$par[9], exp(mle2$par[8])), nrow = 2)

expand.grid(y1 = seq(-5, 5, by = .1), y2 = seq(-5, 5, by = .1)) |>
  rowwise() |>
  mutate(dens1 = hat_p * dmvnorm(c(y1, y2), mean = hat_mu1, sigma = hat_sig1),
         dens2 = (1 - hat_p) * dmvnorm(c(y1, y2), mean = hat_mu2, sigma = hat_sig2),
         dens_mix = dens1 + dens2) -> plot_dens
  
ggplot() +
  geom_contour(aes(y1, y2, z = dens1), colour = "red", data = plot_dens) +
  geom_contour(aes(y1, y2, z = dens2), colour = "blue", data = plot_dens) +
  geom_point(aes(X1, X2, colour = as.character(z)), data = df) +
  geom_point(aes(hat_mu1[1], hat_mu1[2]), shape = 2, colour = "red") +
  geom_point(aes(hat_mu2[1], hat_mu2[2]), shape = 2, colour = "blue") +
  scale_color_manual(values = c("blue", "red")) +
  theme(legend.position = "none")

ggplot() +
  geom_contour(aes(y1, y2, z = dens_mix), data = plot_dens, colour = "black") +
  geom_point(aes(X1, X2), data = df)
```

This seems pretty good... can we break this with initialization?

```{r}
# Centered the second mixture component at a data point, and shrink 
# variance, so normal is super-concentrated around that point.
loglik_mixture(c(.6, 0, 0, 0, 0, y[30, 1], y[30, 2], -50, 0), data = y)  
mle3 <- optim(c(.6, 0, 0, 0, 0, y[30, 1], y[30, 2], -50, 0), loglik_mixture, data = y, control = list(fnscale = -1))
```
```{r, echo = FALSE}
data.frame(Parameter = c("$p$", "$\\mu_{11}$", "$\\mu_{12}$", "$\\Sigma_{111}$", "$\\Sigma_{112}$", "$\\mu_{21}$",  "$\\mu_{22}$", "$\\Sigma_{211}$", "$\\Sigma_{212}$"), 
           Truth = c(p, mu1, sig1[1:2], mu2, sig2[1:2]),
           MLE3 = c(plogis(mle3$par[1]), mle3$par[c(2,3)], exp(mle3$par[4]), mle3$par[5:7], exp(mle3$par[8]), mle3$par[9])) |>
  kable(escape = FALSE, digits = 4)

```


What would change if we were given the complete data, where $Z_i \stackrel{iid}{\sim}\text{Bern}(p)$?

<br/><br/><br/><br/><br/><br/><br/><br/><br/><br/><br/><br/>
<br/><br/><br/><br/><br/><br/>
<br/><br/><br/><br/><br/><br/><br/><br/><br/><br/><br/><br/>
<br/><br/><br/><br/><br/><br/><br/><br/><br/>

Consider the complete log-likelihood:

\begin{align*}
\ell(p, \boldsymbol \mu_1, \boldsymbol \mu_2, \Sigma_1, \Sigma_2 | \boldsymbol Y, \boldsymbol Z) &= \sum\limits_{i = 1}^n\left\{Z_i\log f_1(Y_i; \boldsymbol \mu_1, \Sigma_1) + (1 - Z_i)\log f_2(Y_i; \boldsymbol \mu_2, \Sigma_2) \right. \\
&\qquad \left. + Z_i \log p + (1 - Z_i) \log (1 - p)\right\}.
\end{align*}

We could consider the $Z_i$'s as "weights" which represent our *current* believe in which density each datum come from.

<br/><br/><br/><br/>

Given what our belief is in the weights of the data, what is our estimate of the model parameters?

<br/><br/><br/><br/><br/><br/><br/><br/><br/><br/><br/><br/><br/><br/><br/><br/>
<br/><br/><br/><br/><br/><br/><br/><br/><br/><br/>

This is the basic intuition for the EM algorithm. We will view our data $\boldsymbol Y$ as incomplete and imagine there is missing data $\boldsymbol Z$ that would make the problem simpler if we had it. The EM algorithm then follows:

<br/><br/><br/><br/><br/><br/><br/><br/><br/><br/><br/><br/><br/><br/><br/><br/><br/><br/><br/><br/>

**Example (Two-Component Mixture, Cont'd):** The EM algorithm for the two-component Gaussian mixture model is

<br/><br/><br/><br/><br/><br/><br/><br/><br/>

**Your Turn:** Implement the EM algorithm for the two-component mixture model on our example data.

## Convergence of the EM algorithm

We will show that $\ell\left(\hat{\boldsymbol \theta}^{(k + 1)}\right) \ge \ell\left(\hat{\boldsymbol \theta}^{(k)}\right)$.

<br/><br/><br/><br/><br/><br/>

We know $f_{Z|Y}(z | y; \boldsymbol \theta) = \frac{f_{YZ}(y, z; \boldsymbol \theta)}{f_Y(y | \boldsymbol \theta)}$.

<br/><br/><br/><br/>

Assume we observe $\boldsymbol y = (y_1, \dots, y_n)$, then

<br/><br/><br/><br/><br/><br/><br/><br/>

So, in order to show that $\ell\left(\hat{\boldsymbol \theta}^{(k + 1)}\right) \ge \ell\left(\hat{\boldsymbol \theta}^{(k)}\right)$, this is the same as

<br/><br/><br/><br/><br/><br/><br/><br/>

**Step 1:** Show that $H(\boldsymbol \theta, \boldsymbol \theta^{(k)})$ is maximized when $\boldsymbol \theta = \boldsymbol \theta^{(k)}$.

<br/>

Recall: Jensen's Inequality. A function $\Phi$ is convex if $\Phi(\frac{x_1 + x_2}{2}) \le \frac{1}{2}\Phi(x_1) + \frac{1}{2}\Phi(x_2)$. Then
$$
\Phi(\text{E}[g(X)]) \le \text{E}[\Phi(g(X))],
$$
where $g$ is a real-valued integrable function.

<br/><br/><br/><br/><br/><br/><br/><br/>
<br/><br/><br/><br/><br/><br/><br/><br/>
<br/><br/><br/><br/><br/><br/><br/><br/>
<br/><br/><br/><br/><br/><br/><br/><br/>

**Step 2:** Find a $\boldsymbol \theta^{k + 1}$ that will optimize $Q$.

<br/><br/><br/><br/><br/><br/><br/><br/>
<br/><br/><br/><br/><br/><br/><br/><br/>
<br/><br/><br/><br/><br/><br/><br/><br/>
<br/><br/><br/><br/><br/><br/><br/><br/>
<br/><br/><br/><br/><br/><br/><br/><br/>
<br/><br/><br/><br/><br/><br/><br/><br/>

**Example (Two-Component Mixture, Cont'd):**

<br/><br/><br/><br/><br/><br/><br/><br/>
<br/><br/><br/><br/><br/><br/><br/><br/>
<br/><br/><br/><br/><br/><br/><br/><br/>
<br/><br/><br/><br/><br/><br/><br/><br/>
<br/><br/><br/><br/>

The EM algorithm allows us to obtain $\hat{\boldsymbol \theta}_{\text{EM}}$, the parameter estimate which optimizes the algorithm.

## Variance Estimation for EM estimates

The EM algorithm find the MLE, but it does not automatically produce an estimate of the covariance matrix. Why not?

<br/><br/><br/><br/><br/><br/><br/><br/>

There are several options to estimate the variance.

1. **Bootstrapping**

<br/><br/><br/><br/><br/><br/><br/><br/>
<br/><br/><br/><br/><br/><br/><br/><br/>
<br/><br/><br/><br/><br/><br/><br/><br/>

2. **Louis's Method**

## Another way to cluster: K-means {.page-break-before}

**Goal of clustering:**

<br/><br/><br/><br/><br/><br/><br/><br/>
<br/><br/><br/><br/><br/><br/><br/><br/>

Methods for clustering include hierarchical and non-hierarchical, algorithmic and model-based.

<br/><br/><br/><br/><br/><br/><br/><br/>
<br/><br/><br/><br/><br/><br/><br/><br/>

K-means is a simple and elegant approach to partition a data set into $K$ distinct, non-overlapping clusters.

<br/><br/><br/><br/><br/><br/>

```{r, echo = FALSE, fig.show='hold', out.width="33%", fig.height = 6}
X <- mvtnorm::rmvnorm(100, mean = c(0, 0))
X[1:33,] <- X[1:33,] + matrix(rep(c(-3, -3), 33), ncol = 2, byrow = TRUE)
X[34:67,] <- X[34:67,] + matrix(rep(c(3, 0), 34), ncol = 2, byrow = TRUE)

K2 <- kmeans(X, 2)
K3 <- kmeans(X, 3)
K4 <- kmeans(X, 4)

data.frame(X) |>
  ggplot() +
  geom_point(aes(X1, X2, colour = as.factor(K2$cluster)), size = 2) +
  theme(legend.position = "none", text = element_text(size = 30)) +
  ggtitle("K = 2") +
  xlab("") + ylab("")

data.frame(X) |>
  ggplot() +
  geom_point(aes(X1, X2, colour = as.factor(K3$cluster)), size = 2) +
  theme(legend.position = "none", text = element_text(size = 30)) +
  ggtitle("K = 3") +
  xlab("") + ylab("")

data.frame(X) |>
  ggplot() +
  geom_point(aes(X1, X2, colour = as.factor(K4$cluster)), size = 2) +
  theme(legend.position = "none", text = element_text(size = 30)) +
  ggtitle("K = 4") +
  xlab("") + ylab("")
```

<br/>

The $K$-means clustering procedure results from a simple and intuitive mathematical problem. Let $C_1, \dots, C_K$ denote sets containing the indices of observations in each cluster. These satisfy two properties:

<br/>

1. <br/><br/><br/><br/>

2. <br/><br/><br/><br/>

**Idea:** 

\newpage


The *within-cluster variation* for cluster $C_k$ is a measure of the amount by which the observations within a cluster differ from each other.

<br/><br/><br/><br/><br/><br/>

To solve this, we need to define within-cluster variation.

<br/><br/><br/><br/><br/><br/>

This results in the following optimization problem that defines $K$-means clustering:

<br/><br/><br/><br/><br/><br/>

A very simple algorithm has been shown to find a local optimum to this problem:

\newpage

\newpage

Questions about the algorithm:

1. How do we define distance?

<br/><br/><br/><br/>

2. How do we choose starting values?

<br/><br/><br/><br/>

3. How do we choose $k$?

<br/><br/><br/><br/>

Compared to the Gaussian mixture problem,

# Profile Likelihood {.page-break-before}

